---
title: "Train Accidents"
author: "James Bennett"
date: "8/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 
## Problem Description  

The Federal Rail Road Administration has experience a significant decrease in the frequency of Rail incidents, casualties and deaths over the past years. Despite the reductions in frequency, the relative expense has not been propotionally reduced. In fact, the data shows an increase in accident cost. The average accident cost approximately $281,721,525 per year. The purpose of this report is to develop a better understanding of the contributing factors in severe accident.


## Goal

To identify contributing factors that may help reduce severe train accidents 

## Metrics 

(a) Cost as identified in dollars by the amount of accident damage (ACCDMG), track damage (TRKDMG) and equipment damage (EQPDMG).
(b) Number of casualties as identified by the total number of people killed (TOTKLDS) and injured (TOTINJ) in an accident.



## Data

````{r, include = FALSE}
library(data.table)
library(ggplot2)
library(tidyr)
library(readr)
path <- "/Users/james.bennett/Desktop/Trains/Train_Accidents/Data"

all.files <- list.files(path = path ,pattern = ".txt",full.names = T)

df <- lapply(all.files, fread, sep=",")
dt <- rbindlist( df )
setkey( dt , YEAR, MONTH, DAY, TIMEHR, TIMEMIN )

matrix(names(df[[1]]))
        
source("/Users/james.bennett/Desktop/Trains/Train_Accidents/binder.R")
totacts <- files2DF(df)
`````
````{r, include = FALSE}
h <- colnames(totacts)[-1]
o <- colnames(totacts)[-146]

setnames(totacts, o, h)
````

(1)  Dimensions 

````{r}
dim(totacts)
````
(2)  Most of the accident are minor
````{r}
par(mfcol=c(1,1), oma=c(1,0,0,0), mar=c(1,1,1,0), tcl=-0.1, mgp=c(2,0,0))
boxplot(totacts$ACCDMG, main = "Boxplot of accident damage")

hist(totacts$ACCDMG)
````

(3)  Clean up
(3a) Remove duplicates
````{r}
duplicated(totacts[1:100, c("YEAR", "MONTH", "DAY", "TIMEHR", "TIMEMIN")])

totacts <- totacts[!duplicated(totacts[, c("YEAR", "MONTH", "DAY", "TIMEHR", "TIMEMIN")]),]
````
(3b) Find Columns that have NAs
````{r}

nafind <- function(x){sum(is.na(x))}

apply(totacts,2, "nafind")

# Do we need all the variables?
````
````{r, include = FALSE}
matrix(names(totacts))

# Remove unnecessary variables, then get a summary

nacount <- apply(totacts,2, "nafind")

varWna <- which(nacount > 0)

which(colnames(totacts)[varWna] == "TYPEQ")

varWna <- varWna[-which(colnames(totacts)[varWna]== "TYPEQ")]


#totacts <- totacts[, -varWna]

totacts <- totacts[, -c(5,6,13,26,36,43,45,49,81,85,90,91,92,93,94,95,96,97, 100, 101, 102, 103, 104, 106, 115, 116, 117, 118, 121, 137, 140, 141, 142, 143, 144, 145, 146)]
write.csv(totacts, file ="/Users/james.bennett/Desktop/Trains/Output/totactsClean.csv", row.names = F)
````
(3c)  Summary of Clean up 

````{r}
dim(totacts) # How Many Accidents
````
## Analysis
The total cost from **2002 to 2019** 
````{r}

sum(totacts$ACCDMG)  ###Total cost of Accidents is $5,464,913,062
````
The average cost per year is 
````{r}

sum(totacts$ACCDMG)/18  ### Average cost per year is 303,606,281
dmgyrsum <- tapply(totacts$ACCDMG, totacts$YEAR, sum)
sum(totacts$TOTKLD) # Total number killed - 819 People


max(totacts$TOTKLD) # Largest number killed in an accident - 25 people


sum(totacts$TOTINJ) # Total number injured - 8216 Injuries


max(totacts$TOTINJ) ##what is the most people injured in an accident 1000

round(sum(totacts$TOTINJ)/18) # What is the average number of injuries per year?  456
````

````{r, include = FALSE}
dmgyrsum <- tapply(totacts$ACCDMG, totacts$YEAR, sum)
````
## Time Series of Accidents
````{r}
#**************************************************
#
#   Time series of Accidents
#
#**************************************************

# Yearly no. of accidents

plot(2:max(totacts$YEAR), tapply(totacts$ACCDMG, totacts$YEAR, length), type = "l", col = "black", xlab = "Year", ylab = "Frequency", main = "Number of Accidents per Year", lwd =2)


# Yearly total cost of accidents

plot(2:max(totacts$YEAR), tapply(totacts$ACCDMG, totacts$YEAR, sum), type = "l", col = "black", xlab = "Year", ylab = "Cost ($)", main = "Total Damage per Year", lwd =2)

symbols(2002:2019, tapply(totacts$ACCDMG, totacts$YEAR, sum), circles=tapply(totacts$ACCDMG, totacts$YEAR, max),inches=0.35, fg="white", bg="red", xlab="Year", ylab="Cost ($)", main = "Total Accident Damage")
lines(2002:2019, tapply(totacts$ACCDMG, totacts$YEAR, sum))
# Yearly maximum cost of accidents

plot(2:max(totacts$YEAR), tapply(totacts$ACCDMG, totacts$YEAR, max), type = "l", col = "black", xlab = "Year", ylab = "Cost ($)", main = "Max Damage per Year", lwd =2)

symbols(2002:2019, tapply(totacts$ACCDMG, totacts$YEAR, max), circles=tapply(totacts$ACCDMG, totacts$YEAR, max),inches=0.35, fg="white", bg="red", xlab="Year", ylab="Cost ($)", main = "Max Accident Damage per Year")
lines(2002:2019, tapply(totacts$ACCDMG, totacts$YEAR, max))



# Total killed and total injured and the sum of them.
plot(2:max(totacts$YEAR), tapply(totacts$TOTKLD, totacts$YEAR, sum), type = "l", col = "black", xlab = "Year", ylab = "Cost ($)", main = "Total Deaths per Year", lwd =2)

symbols(2002:2019, tapply(totacts$TOTKLD, totacts$YEAR, sum), circles = tapply(totacts$TOTKLD, totacts$YEAR, max), inches = 0.35, fg ="yellow", bg ="red", xlab = "Year", ylab = "People Killed", main = "Total Killed")
lines(2002:2019, tapply(totacts$TOTKLD, totacts$YEAR, sum))

plot(2:max(totacts$YEAR), tapply(totacts$TOTKLD, totacts$YEAR, max), type = "l", col = "black", xlab = "Year", ylab = "Cost ($)", main = "Max Deaths per Year", lwd =2)

symbols(2002:2019, tapply(totacts$TOTKLD, totacts$YEAR, max), circles = tapply(totacts$TOTKLD, totacts$YEAR, max), inches = 0.35, fg ="yellow", bg ="red", xlab = "Year", ylab = "People Killed", main = "Max Killed per Year")
lines(2002:2019, tapply(totacts$TOTKLD, totacts$YEAR, max))

plot(2:max(totacts$YEAR), tapply(totacts$TOTKLD, totacts$YEAR, sum), type = "l", col = "black", xlab = "Year", ylab = "Cost ($)", main = "Total Injuries per Year", lwd =2)

symbols(2002:2019, tapply(totacts$TOTINJ, totacts$YEAR, sum), circles = tapply(totacts$TOTINJ, totacts$YEAR, max), inches = 0.35, fg ="black", bg ="red", xlab = "Year", ylab = "People Injured", main = "Total Injured")
lines(2002:2019, tapply(totacts$TOTINJ, totacts$YEAR, sum))

plot(2:max(totacts$YEAR), tapply(totacts$TOTKLD, totacts$YEAR, max), type = "l", col = "black", xlab = "Year", ylab = "Cost ($)", main = "Max Injuries per Year", lwd =2)

symbols(2002:2019, tapply(totacts$TOTINJ, totacts$YEAR, max), circles = tapply(totacts$TOTINJ, totacts$YEAR, max), inches = 0.35, fg ="black", bg ="red", xlab = "Year", ylab = "People Injured", main = "Max Injured per Year")
lines(2002:2019, tapply(totacts$TOTINJ, totacts$YEAR, max))
````

## Accidents and the weather
        We can see that accidents are normally distributed accross the temperature scale
````{r}
par(mfrow = c(2,2))

hist(totacts$TEMP, breaks = "scott", main = "Accident Temperatures (Scott)", xlab = "Temp (F)", col = "steelblue")

hist(totacts$TEMP, breaks = "fd", main = "Accident Temperatures (FD)", xlab = "Temp (F)", col = "steelblue")

hist(totacts$TEMP, main = "Accident Temperatures (Sturges)", xlab = "Temp (F)", col = "steelblue")

hist(totacts$TEMP, breaks = 100, main = "Accident Temperatures (100)", xlab = "Temp (F)", col = "steelblue")

par(mfrow = c(1,1))
````

## Barplots of Metrics and Extremes

````{r}
boxplot(totacts$ACCDMG, main = "Xtreme Accident damage")
boxplot(totacts$TOTKLD, main = "Deaths in Extreme incident")
````
````{r, include = FALSE}
dmgbox <- boxplot(totacts$ACCDMG)

dmgbox2 <- boxplot(totacts$TOTKLD)

# How many extreme damage accidents?

length(dmgbox$out) ##extreme accident dmg 5922

length(dmgbox2$out) ## 631 accidents involving deaths (this is not commom)
dmgbox$stats
````
````{r}

# What proportion of accidents are extreme? (round to 2 digits) - 13%
round(length(dmgbox$out)/length(totacts$ACCDMG),2)

# What is the proportion of costs for extreme damage accidents? (round to 2 digits)
round(sum(dmgbox$out)/sum(totacts$ACCDMG),2) ##13% causes 74% of the damages - Insanity!!
# Create a data frame with just the extreme ACCDMG accidents
round(length(dmgbox2$out)/length(totacts$TOTKLD),2)
##.01 are extreme - deaths are an ware event
round(sum(dmgbox2$out)/sum(totacts$TOTKLD),2)

##all deaths are were events

xdmg <- totacts[totacts$ACCDMG > dmgbox$stats[5],]

dim(xdmg)
###4862 are were 

# Look at the boxplots and histograms of these extreme accidents

boxplot(xdmg$ACCDMG, col = "steelblue", main = "Accidents with Extreme Damage", ylab = "Cost ($)")

plot(1:18, tapply(xdmg$ACCDMG, xdmg$YEAR, sum), type = "l", xlab = "Year", ylab = "Total Damage ($)", main = "Total Extreme Accident Damage per Year")

# also plot number of accidents per year.

plot(1:18, tapply(xdmg$ACCDMG, xdmg$YEAR, length), type = "l", xlab = "Year", ylab = "No. of Accidents", main = "Number of Extreme Accidents per Year")

# Frequency of accident types

barplot(table(xdmg$TYPE)) #compare with the totacts plot
##Lots of Derailments - wonder is speeding has to do with this - Type = 1

# Repeat for TOTKLD and TOTINJ
# Create a variable called Casualty = TOTKLD + TOTINJ

max(totacts$TOTINJ) ##1000 in a single accident 
max(totacts$TOTKLD) ##25 in a single maccident
Casualidad = totacts$TOTKLD + totacts$TOTINJ
max(Casualidad) ###1001
plot(2:max(totacts$YEAR), tapply(totacts$TOTKLD, totacts$YEAR, max), type = "l", col = "black", xlab = "Year", ylab = "Frequency", main = "Number of KILLED", lwd =2)
plot(2:max(totacts$YEAR), tapply(totacts$TOTINJ, totacts$YEAR, max), type = "l", col = "black", xlab = "Year", ylab = "Frequency", main = "Number of Injured", lwd =2)
plot(2:max(totacts$YEAR), tapply(Casualidad, totacts$YEAR, max), type = "l", col = "blue", xlab = "Year", ylab = "Frequency", main = "Combined Casualties", lwd =2)
````
